GPU-friendly Stroke Expansion

### 2024-06-14 paper.tex
More low-hanging fixes
- Add non-breaking space before citenum
- Use "ms" instead of "fps" numbers in renderings figure caption

We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins. The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth. As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints. Standards for graphics formats, such as SVG[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior. While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.

A great number of rendering techniques for strokes have been proposed. Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke. Included in this category is polar stroking[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error. Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering. Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \citet{Nehab2020}. In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \citet{Nehab2020}, their evolutes when targeting strong correctness. For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \citet{Nehab2020}.

The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}. A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.

@@

In the principled, correct specification for stroking[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width. When it does, additional segments must be drawn, including evolutes of the original curve. In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques. By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve[\citenum{Yoshida2012}].

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.

@@

An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm. For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}[\citenum{Laine2011}] and \emph{MPVG}[\citenum{Ganacim2014}]. In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles. The final compute stage of our renderer (called \emph{fine rasterizer}) operates at tile granularity and computes pixel coverage for filled polygons outlined by the line soup (Figure~\ref{fig:pipeline} illustrates the overall structure of our renderer, the details of which are beyond the scope of this paper).

@@

We authored two additional test scenes in order stress the GPU with a higher workload. The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus. We used two variants of this scene with \emph{butt} and \emph{round} cap styles. For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths. We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively. Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.

@@

We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment. We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}). For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth. We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.

@@

\begin{figure*}
    \centering
    \includegraphics[scale=0.2]{test_renderings}
    \caption{Renderings of some of our test scenes created using our algorithm and associated frame rates. All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline. \textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
    \Description{Renderings of the test scenes}
    \label{fig:renderings}
\end{figure*}

@@

\begin{itemize}
    \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be. An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph[\citenum{Patel2024}]. This structure is appealing because load balancing is done by hardware.

>>>>

We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins. The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth. As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints. Standards for graphics formats, such as SVG~[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior. While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.

A great number of rendering techniques for strokes have been proposed. Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke. Included in this category is polar stroking~[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error. Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering. Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \citet{Nehab2020}. In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \citet{Nehab2020}, their evolutes when targeting strong correctness. For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \citet{Nehab2020}.

The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson~[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}. A variant is also used in Skia~[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago~[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.

@@

In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width. When it does, additional segments must be drawn, including evolutes of the original curve. In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques. By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve~[\citenum{Yoshida2012}].

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm~[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.

@@

An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm. For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}~[\citenum{Laine2011}] and \emph{MPVG}~[\citenum{Ganacim2014}]. In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles. The final compute stage of our renderer (called \emph{fine rasterizer}) operates at tile granularity and computes pixel coverage for filled polygons outlined by the line soup (Figure~\ref{fig:pipeline} illustrates the overall structure of our renderer, the details of which are beyond the scope of this paper).

@@

We authored two additional test scenes in order stress the GPU with a higher workload. The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus. We used two variants of this scene with \emph{butt} and \emph{round} cap styles. For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark~[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths. We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively. Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.

@@

We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment. We found that Wang's formula~[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}). For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth. We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.

@@

\begin{figure*}
    \centering
    \includegraphics[scale=0.2]{test_renderings}
    \caption{Renderings of some of our test scenes created using our algorithm. All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline. \textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (2.94~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (13.89~ms); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (25~ms including CPU-side dashing - the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.05~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.23~ms).}
    \Description{Renderings of the test scenes}
    \label{fig:renderings}
\end{figure*}

@@

\begin{itemize}
    \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be. An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph~[\citenum{Patel2024}]. This structure is appealing because load balancing is done by hardware.
